---
title: "Variant calling"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Variant calling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css echo=FALSE}
img {
    border: 0px !important;
    margin: 2em 2em 2em 2em !important;
}
code {
    border: 0px !important;
}
```

```{r echo=FALSE, results="hide"}
knitr::opts_chunk$set(
    cache = FALSE,
    echo = TRUE,
    collapse = TRUE,
    comment = "#>"
)
suppressPackageStartupMessages(library(pepitope))
```

This vignette describes how to go from variants (and optionally gene fusions) to
the Minigene Library report.

We will usually work with standard processing pipelines for variant calling,
RNA-seq quantification, and RNA fusion calling. The best-known collection of
these workflows is [NF-core](https://nf-co.re/).

It provides the following pipelines for these tasks:

* [`sarek`](https://nf-co.re/sarek/) for variant calling using HaploytypeCaller and Mutect2
* [`rnaseq`](https://nf-co.re/rnaseq/) for RNA-seq quantification using STAR
* [`rnafusion`](https://nf-co.re/rnafusion/) for RNA fusion quantification using multiple algorithms

Running these will usually require access to high performance computing
facilities, and you already get the processed files from your core facility
or sequencing provider.

```{r}
library(pepitope)

# SNPs and small indels, e.g. from 'sarek' nf-core pipeline
variant_vcf_file = "my_variants.vcf.gz"

# RNA counts and TPM file, e.g. from 'rnaseq' nf-core pipeline
rna_counts_file = "count_file.tsv"
rna_tpm_file = "tpm_file.tsv"

# Combined fusion VCF file, e.g. from 'rnafusion' nf-core pipeline
fusion_vcf_file = "my_fusions.vcf.gz"
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# set up all the actually required files in the background
tdir = tempdir()
variant_vcf_file = file.path(tdir, variant_vcf_file)
vars = sim_variants()
VariantAnnotation::writeVcf(vars, variant_vcf_file)
```

Prerequisites
-------------

### Selecting the right reference genome

The variant calling and RNA-seq counts were mapped to a reference genome and
gene annotations. It is important to keep these consistent between the NF-core
processing pipelines and the Minigene Library annotation.

We are usually working with `BSgenome` (for the reference genome) and `EnsDb`
(for the gene annotations) objects. For human data, the most widely used
reference genome is `GRCh38`, and a recent `Ensembl` annotation release.

With our test data, we know that `GRCh38` is the correct reference genome and
`Ensembl 106` is the correct version of gene annotations. We can get both
objects from the `BSgenome` and `AnnotationHub` Bioconductor packages, respectively.

```{r}
ens106 = AnnotationHub::AnnotationHub()[["AH100643"]]
asm = BSgenome.Hsapiens.NCBI.GRCh38::BSgenome.Hsapiens.NCBI.GRCh38
# rna_counts = readr::read_tsv(rna_counts_file)
```

A caveat here is that the chromosome prefixes need to be consistent between the
variants in the VCF file and the genome/gene annotations. There are two
"styles", either UCSC (includes "chr" prefix) or NCBI/Ensembl (without "chr"
prefix).

The `sarek` pipline uses UCSC prefixes on the GRCh38 genome, so we need to switch
the genome and gene annotation styles:

```{r}
seqlevelsStyle(ens106) = "UCSC"
seqlevelsStyle(asm) = "UCSC"
asm@seqinfo@genome[] = "GRCh38"
```

The correct styles will depend on how your VCF files were generated.

### Adding RNA expression

The NF-core `rnaseq` workflow will provide two gene expression files, one for
raw read counts and one for transcripts per million (TPM). These contain all
samples in a run, so we need to subset them to the current sample we are
interested in. These files are usually called:

* `salmon.merged.gene_counts.tsv`
* `salmon.merged.gene_tpm.tsv`

We can combine and subset them the following way:

```{r eval=FALSE}
rna_counts = readr::read_tsv(rna_counts_file)
rna_tpm = readr::read_tsv(rna_tpm_file)

counts = readr::read_tsv(args$rna_count) |> dplyr::select(gene_id, gene_name, count=SAMPLE)
tpm = readr::read_tsv(args$rna_tpm) |> dplyr::select(gene_id, gene_name, tpm=SAMPLE)

rna_sample = inner_join(counts, tpm)
```

__ SHOW TABLE HERE __

SNPs and small indels
---------------------

### Reading and filtering mutations

```{r eval=FALSE}
vr1 = readVcfAsVRanges(variant_vcf_file) |>
    filter_variants(min_cov=2, min_af=0.05, pass=TRUE)
```
```{r echo=FALSE}
# pass does not work with example data
vr1 = readVcfAsVRanges(variant_vcf_file)
genome(vr1) = "GRCh38" # save/load error
vr1$AF = vars$AF # does not roundtrip via saved VCF
vr1 = filter_variants(vr1, min_cov=2, min_af=0.05)
fus = DataFrame() # we don't have simulated fusions yet
```

Here, we are using the following filters:

* `min_cov=2` -- a variant needs to be covered by at least 2 reads
* `min_af=0.05` -- a variant needs to occur in at leat 5% of reads
* `pass=TRUE` -- a variant needs to pass the standard QC filters

__ SHOW TABLE HERE __

### Annotating and subsetting expressed variants

```{r warning=FALSE}
ann = annotate_coding(vr1, ens106, asm)
subs = ann |>
#    filter_expressed(rna_sample, min_reads=1, min_tpm=0) |>
    subset_context(15)
```

Here, we are using the following filters:

* `min_reads=1` -- the gene needs to have at least one RNA read
* `min_tpm=0` -- we do not apply an additional TPM filter

In addition, we set the region of interest (context) to 15 codons up- and downstream of the
variant. Hence, a SNP will have a total length of 93 nucleotides (15*3 + the SNP codon itself).
An insertion will have the inserted sequence and 15 codons, a deletion only 15 codons both
sides. A frameshift will have 15 codons upstream and the entire sequence downstream until a
STOP codon is reached. The latter may extend into the 3' UTR.

__ SHOW TABLE HERE __

### Phasing with germline variants

```{r}
# not implemented yet
```

Fusion genes from RNA-seq
-------------------------

### Annotating fusion genes

```{r eval=FALSE}
vr2 = readVcfAsVRanges(fusion_vcf_file) |>
    filter_fusions(min_reads=2, min_split_reads=1, min_tools=1)

seqlevelsStyle(vr2) = "UCSC"
```

Here, we are using the following filters:

* `min_reads=2` -- the fusion needs to be supported by 2 split or pair distance reads
* `min_split_reads=1` -- the fusion needs to be supported by at least one split read
* `min_tools=1` -- the fusion needs to be reported by at least one tool

__ SHOW TABLE HERE __

### Subsetting expressed fusion genes

```{r eval=FALSE}
fus = annotate_fusions(vr2, ens106, asm) |>
    filter_expressed(rna_sample, min_reads=1, min_tpm=0) |>
    subset_context(15)
```

Here, we are using the following filters:

* `min_reads=1` -- the gene needs to have at least one RNA read
* `min_tpm=0` -- we do not apply an additional TPM filter

In addition, we set the region of interest to 15 codons up- and downstream of the fusion site.

__ SHOW TABLE HERE __

Generating the Minigene Library
-------------------------------

### Tiling cDNAs of interest into smaller peptides

```{r}
tiled = pep_tile(subs) |>
    remove_cutsite(BbsI="GAAGAC")
```

__ SHOW TABLE HERE __

### Adding barcodes

```{r}
# this is not yet implemented
```

__ SHOW TABLE HERE __

### Saving a report file

We can then combine our generated tables into a report save it with the
`writexl` package:

```{r eval=FALSE}
report = make_report(ann, subs, fus, tiled)
writexl::write_xlsx(report, "report_file.xlsx")
```

This `.xlsx` file will contain the different tables as sheets. We will use it 
as an annotation file in the quality control and screen steps.
